Dynamical modeling of tidal streams

### 2013-12-24 tex/ms.tex
streams: re-arrange figures
streams: first pass of the appendix

\renewcommand{\vec}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\unitvec}[1]{\ensuremath{\mathbf{\hat{#1}}}}
\newcommand{\vecx}{\ensuremath{\vec{x}}}
\newcommand{\vecv}{\ensuremath{\vec{v}}}
\newcommand{\vech}{\ensuremath{\vec{H}}}
\newcommand{\vecj}{\ensuremath{\vec{J}}}
\newcommand{\vecp}{\ensuremath{\vec{p}}}
\newcommand{\veco}{\ensuremath{\vec{\Omega}}}
\newcommand{\veca}{\ensuremath{\boldsymbol\theta}}
\newcommand{\df}{\ensuremath{f}}
\newcommand{\tdf}{\ensuremath{\mathrm{tDF}}}
\newcommand{\paramsdiff}{\ensuremath{\vec{p}_\Delta}}
\newcommand{\paramsdf}{\ensuremath{\vec{p}_{\mathrm{DF}}}}
\newcommand{\paramspot}{\ensuremath{\vec{p}_\Phi}}

\newcommand{\DF}{\df}
\newcommand{\dff}{\ensuremath{\mathbf{f}}}
\newcommand{\dex}{\ensuremath{\,\mathrm{dex}}}
\newcommand{\Gyr}{\ensuremath{\,\mathrm{Gyr}}}
\newcommand{\kpc}{\ensuremath{\,\mathrm{kpc}}}
\newcommand{\pc}{\ensuremath{\,\mathrm{pc}}}
\newcommand{\kms}{\ensuremath{\,\mathrm{km\ s}^{-1}}}
\newcommand{\msun}{\ensuremath{\,\mathrm{M}_{\odot}}}
\newcommand{\inv}{\ensuremath{^{-1}}}

@@

\acknowledgements It is a pleasure to thank \dots for helpful comments
and assistance. J.B. was supported by NASA through Hubble Fellowship
grant HST-HF-51285.01 from the Space Telescope Science Institute,
which is operated by the Association of Universities for Research in
Astronomy, Incorporated, under NASA contract NAS5-26555. J.B.
%and H.-W.R  (BOVY: ALSO ACKNOWLEDGE WITHOUT S)
acknowledges support from SFB 881 (A3) funded by the German
Research Foundation DFG.

@@

\section{Dynamical modeling of two-star streams}\label{sec:twostar}

@@

\bibitem[{{Binney} \& {Tremaine}(2008)}]{binneytremaine}
  Binney,~J. \& Tremaine,~S. 2008, Galactic Dynamics: Second Edition
\bibitem[Binney(2008)]{Binney08a}
  Binney,~J. 2008, \mnras, 386, L47
\bibitem[Carlberg(2009)]{Carlberg09a}
  Carlberg,~R.~G. 2009, \apj, 705, L223
\bibitem[Carlberg(2012)]{Carlberg12a}
Carlberg,~R.~G. 2012, \apj, 748, 20
\bibitem[Eyre \& Binney(2011)]{Eyre11a}
  Eyre,~A. \& Binney,~J. 2011, \mnras, 413, 1852
\bibitem[Helmi \& White(1999)]{Helmi99a}
  Helmi,~A. \& White,~S.~D.~M. 1999, \mnras, 307, 495
\bibitem[Johnston \etal(1999)]{Johnston99a}
  Johnston,~K.~V., Zhao,~H., Spergel,~D.~N., \& Hernquist,~L. 1999, \apj, 512, L109
\bibitem[Koposov \etal(2010)]{Koposov10a}
  Koposov,~S.~E., Rix,~H.-W., \& Hogg,~D.~W. 2010, \apj, 712, 260
\bibitem[Ngan \& Carlberg(2014)]{Ngan14a}
  Ngan,~W.~H.~W. \& Carlberg,~R.~G. 2014, \apj, submitted
\bibitem[Pe{\~n}arrubia \etal(2012)]{Penarrubia12a}
  Pe{\~n}arrubia, J., Koposov,~S.~E., \& Walker,~M.~G. 2012, \apj, 760, 2
\bibitem[Price-Whelan \& Johnston(2013)]{PriceWhelan13a}
  Price-Whelan,~A.~M. \& Johnston,~K.~V.~2013, \apj, 778, L12
\bibitem[Sanders(2012)]{Sanders12a}
  Sanders,~J. 2012, \mnras, 426, 128
\bibitem[Sanders \& Binney(2013a)]{Sanders13a}
  Sanders,~J. \& Binney,~J. 2013a, \mnras, 433, 1813
\bibitem[Sanders \& Binney(2013b)]{Sanders13b}
  Sanders,~J. \& Binney,~J. 2013b, \mnras, 433, 1826
\bibitem[Tremaine(1999)]{Tremaine99a}
  Tremaine,~S. 1999, \mnras, 307, 877
\bibitem[Yoon \etal(2011)]{Yoon11a}
  Yoon,~J.~H., Johnston,~K.~V., \& Hogg,~D.~W. 2011, \apj, 731, 58
\end{thebibliography}

@@

>>>>

\renewcommand{\vec}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\unitvec}[1]{\ensuremath{\mathbf{\hat{#1}}}}
\newcommand{\vecx}{\ensuremath{\vec{x}}}
\newcommand{\vecv}{\ensuremath{\vec{v}}}
\newcommand{\vech}{\ensuremath{\vec{H}}}
\newcommand{\vecj}{\ensuremath{\vec{J}}}
\newcommand{\vecp}{\ensuremath{\vec{p}}}
\newcommand{\vecn}{\ensuremath{\vec{n}}}
\newcommand{\veco}{\ensuremath{\vec{\Omega}}}
\newcommand{\veca}{\ensuremath{\boldsymbol\theta}}
\newcommand{\df}{\ensuremath{f}}
\newcommand{\tdf}{\ensuremath{\mathrm{tDF}}}
\newcommand{\paramsdiff}{\ensuremath{\vec{p}_\Delta}}
\newcommand{\paramsdf}{\ensuremath{\vec{p}_{\mathrm{DF}}}}
\newcommand{\paramspot}{\ensuremath{\vec{p}_\Phi}}

\newcommand{\DF}{\df}
\newcommand{\dff}{\ensuremath{\mathbf{f}}}
\newcommand{\dex}{\ensuremath{\,\mathrm{dex}}}
\newcommand{\Myr}{\ensuremath{\,\mathrm{Myr}}}
\newcommand{\Gyr}{\ensuremath{\,\mathrm{Gyr}}}
\newcommand{\kpc}{\ensuremath{\,\mathrm{kpc}}}
\newcommand{\pc}{\ensuremath{\,\mathrm{pc}}}
\newcommand{\kms}{\ensuremath{\,\mathrm{km\ s}^{-1}}}
\newcommand{\msun}{\ensuremath{\,\mathrm{M}_{\odot}}}
\newcommand{\inv}{\ensuremath{^{-1}}}

@@

\acknowledgements It is a pleasure to thank \dots for helpful comments
and assistance. J.B. was supported by NASA through Hubble Fellowship
grant HST-HF-51285.01 from the Space Telescope Science Institute,
which is operated by the Association of Universities for Research in
Astronomy, Incorporated, under NASA contract NAS5-26555. J.B.
acknowledges support from SFB 881 (A3) funded by the German Research
Foundation DFG.

@@

The wide-spread use of action--angle coordinates, in many ways the
natural coordinates for studying the orbital structure of galaxies, in
dynamical modeling has been frustrated by the difficulty in
calculating the transformation to and from these coordinates,
$(\vecx,\vecv) \leftrightarrow (\vecj,\veca)$. The most general method
for calculating the transformation $(\vecj,\veca) \rightarrow
(\vecx,\vecv)$ is provided by torus modeling \citep{McGill90a}. In the
context of modeling observational data, computing the transformation
$(\vecx,\vecv) \rightarrow (\vecj,\veca)$ is more important and the
most general method consists of iteratively inverting the torus
modeling \citep{McMillan08a}, which is computationally
expensive. Recently, progress has been made in computing
$(\vecx,\vecv) \rightarrow (\vecj,\veca)$ for Milky-Way-like
potentials by either implicitly or explicitly approximating the
gravitational potential as a St\"{a}ckel potential
\citep{Binney12a,Sanders12a}, which allows action--angle coordinates
to be computed with percent-level errors for moderately eccentric
orbits. However, these methods break down for orbits with radial
and/or vertical actions of a similar magnitude as the angular
momentum---that is, orbits that feel the gravitational potential over
a volume large enough that the St\"{a}ckel approximation fails. This
is especially problematic for the orbits of stars in tidal streams, as
the progenitors of tidal streams are typically on quite eccentric
orbits \citep[\eg,][]{Sanders13a}.

Recently, Fox \& Binney, 2014, in preparation, have suggested a new
method for calculating $\vecj(\vecx,\vecv)$ inspired by torus
modeling. This method requires only a single orbit integration and
allows the actions to be calculated accurately and as precisely as
desired (that is, a convergence criterion can be applied and
convergence can be attained). I describe this method briefly here and
then show how to extend it to also calculate the frequencies and
angles: $(\vecx,\vecv) \rightarrow (\veco,\veca)$.

As in torus modeling, the Fox \& Binney method uses a auxiliary
isochrone potential $\Phi^A$ that should be close to the target
potential $\Phi$ for which the action--angle coordinates should be
calculated (a precise definition of ``close'' will be given
below). Actions and angles $(\vecj^A,\veca^A)$ calculated in the
auxiliary potential can be related to those in the target potential
$(\vecj,\veca)$ by a generating function that can be written as
\citep{McGill90a}
\begin{equation}
  S(\veca^A,\vecj) = \veca^A\cdot\vecj+2\sum_{\vecn > 0} S_{\vecn}(\vecj)\sin(\vecn\cdot\veca^A)\,,
\end{equation}
where the $\vecn > 0$ sum is over the integer two-dimensional
half-plane that excludes the origin (two- rather than
three-dimensional, as the azimuthal angle is perfectly correlated with
the vertical angle in the auxiliary isochrone potential;
\citealt{binneytremaine}), and the $S_{\vecn}$ are functions of the
target actions that are to be determined. The canonical transformation
corresponding to this generating function is
\begin{equation}\label{eq:jjt}
  \vecj^A = \frac{\partial S(\veca^A,\vecj)}{\partial \veca^A} = \vecj + 2\sum_{\vecn > 0} \vecn\,S_{\vecn}(\vecj)\cos(\vecn\cdot\veca^A)\,,
\end{equation}
and
\begin{equation}\label{eq:aat}
  \veca = \frac{\partial S(\veca^A,\vecj)}{\partial \vecj} = \veca^A +2\sum_{\vecn > 0} \frac{\partial S_{\vecn}(\vecj)}{\partial \vecj}\,\sin(\vecn\cdot\veca^A)\,.
\end{equation}

The Fox \& Binney method consists of averaging \eqnname~(\ref{eq:jjt})
over the auxiliary angle along the path of the orbit, such that the
oscillating cosine behavior cancels and the target action is obtained
as the auxiliary-angle-averaged auxilary action. That is,
\begin{equation}\label{eq:jjtint}
  \int \dd \theta_i^A J_i^A = \int \dd \theta_i^A J_i + 2 \sum_{\vecn > 0} \vecn\,S_{\vecn}(J_i)\int \dd \theta_i^A\cos(\vecn\cdot\veca^A)\,,
\end{equation}
where the integral is over auxiliary angles calculated on the path of
the orbit in the target potential. The factor $\vecn\,S_{\vecn}(J_i)$
comes out of the integral as the target actions are conserved along
the orbit. In this Equation, we integrate over the angle coordinate
$\theta_i^A$ associated with the action $J_i$ that we want to compute
(that is, $\theta_R$ for $J_R$ and so on). This is because of the
factor $\vecn$: the condition $\vecn > 0$ means that when calculating
$J_R$ there are terms with $\vecn = (n_R,0)$, which would not cancel
according to the argument below when integrating over $\theta_Z$. For
non-resonant orbits and after a long-enough orbit integration, the
angles can be considered to be independent (that is, the orbit ``fills
its torus''), such that we can write (for example, for $\theta_R$)
\begin{equation}
  \int \dd \theta_R^A\cos(\vecn\cdot\veca^A) = \mathrm{Re}\left[\int \dd \theta_R^A\,\exp(-i\vecn\cdot\veca^A)\right] = \mathrm{Re}\left[\exp(-i\,n_Z\,\theta^A_Z)\int \dd \theta_R^A\,\exp(-i\,n_R\,\theta^A_R)\right]\,.
\end{equation}
As long as the auxiliary angle $\theta_R$ goes through the full range
$0$ to $2\pi$ along the orbit, the oscillatory behavior integrates to
zero and \eqnname~(\ref{eq:jjtint}) simplifies to
\begin{equation}\label{eq:aAIJ}
  J_i = \frac{\int \dd \theta_i^A J_i^A}{\int \dd \theta_i^A}\,,
\end{equation}
that is, the target action is obtained by averaging the auxiliary
action over the auxiliary angle.

The left panel of \figurename~\ref{fig:aAI} shows a running average of
the radial action calculated using \eqnname~(\ref{eq:aAIJ}) along the
orbit of the progenitor in the simulation used in the main body of
this paper. The time in this figure is in units of $35.56\Myr$, so the
orbit is integrated for a total of $7.11\Gyr$. After a few radial
periods, the radial action quickly converges with a remaining
oscillatory behavior that is primarily a function of the radial angle,
but that becomes smaller over time. With longer integrations and
shorter time-steps, the radial action can converge to any desired
tolerance. The duration of the orbit integration shown in this figure
is what is typically used in this paper and gives a precision of
$\approx2\,\%$ in the actions.

Given the orbit integration already performed for calculating the
actions, we can use \equationname~(\ref{eq:aat}) to calculate the
frequencies and angles corresponding to the point $(\vecx,\vecv)$
(this procedure is similar to that used in \citealt{McMillan08a}). We
first re-write \equationname~(\ref{eq:aat}) in terms of the angle at
time zero and the frequencies
\begin{equation}\label{eq:aat2}
  \veca = \veca(t=0) + \veco(\vecj) \,t = \veca^A +2\sum_{\vecn > 0} \frac{\partial S_{\vecn}(\vecj)}{\partial \vecj}\,\sin(\vecn\cdot\veca^A)\,.
\end{equation}
This a linear system of equations for each point along the orbit with
unknowns $\veca(t=0)$, $\veco(\vecj)$, and the functions
$\frac{\partial S_{\vecn}(\vecj)}{\partial \vecj}$. For a finely
integrated orbit, this system is highly over-constrained (at least
when the number of expansion terms is small) and its solution can be
found by simple linear algebra for any desired number of expansion
terms $\sin(\vecn\cdot\veca^A)$. For extra stability, we integrate the
orbit both forward and backward in time, such that the desired angle
$\veca(t=0)$ is at the center of the dependent variable $t$.

The right panel of \figurename~\ref{fig:aAI} shows the auxiliary
radial angle versus the auxiliary vertical angle along the orbit for
the progenitor used in the main body of this paper. The periodic
behavior in the angles is removed before the linear fit described in
the previous paragraph, which is shown in this figure. It is clear
that the contribution of the oscillatory terms
$\sin(\vecn\cdot\veca^A)$ is small, such that a small number of
expansion terms is sufficient (we limit the terms to $n_R < 4$ and
$n_Z < 4$).

The method described here is completely general, but the only obstacle
to using it is that a sufficiently close isochrone potential has to be
supplied. Sufficiently close means specifically that both the
auxiliary radial and vertical angle have to go through the full range
of $0$ to $2\pi$ along the orbit in the target potential. Whether they
do or not is easy to check and I have found that with limited trial
and error a good enough isochrone potential can be determined quickly
for Milky-Way-like axisymmetric potentials, although it remains to be
checked whether this holds for triaxial potentials as well. For
analyzing stream data, the necessity of this trial-and-error
determinition of the auxiliary potential does not pose a practical
problem, as all of the stream stars are on very similar
orbits. Therefore, the same auxiliary potential can be used for all
action--angle calculations for a given potential and for different
potentials as well, as long as they are not too different. More
widespread use of this technique may require automated methods for
determining a good auxiliary potential. In the unlikely circumstance
that no good auxiliary isochrone potential can be found, the method
discussed here can also be used with other auxiliary potentials for
which the actions and angles can be calculated, such as other
spherical potentials or the family of St\"{a}ckel potentials. This
will, however, add a non-negligible amount of computation for the
necessary numerical integrations.

@@

\bibitem[{{Binney} \& {Tremaine}(2008)}]{binneytremaine}
  Binney,~J. \& Tremaine,~S. 2008, Galactic Dynamics: Second Edition
\bibitem[Binney(2008)]{Binney08a}
  Binney,~J. 2008, \mnras, 386, L47
\bibitem[Binney(2012)]{Binney12a}
  Binney,~J. 2012, \mnras, 426, 1324
\bibitem[Carlberg(2009)]{Carlberg09a}
  Carlberg,~R.~G. 2009, \apj, 705, L223
\bibitem[Carlberg(2012)]{Carlberg12a}
Carlberg,~R.~G. 2012, \apj, 748, 20
\bibitem[Eyre \& Binney(2011)]{Eyre11a}
  Eyre,~A. \& Binney,~J. 2011, \mnras, 413, 1852
\bibitem[Helmi \& White(1999)]{Helmi99a}
  Helmi,~A. \& White,~S.~D.~M. 1999, \mnras, 307, 495
\bibitem[Johnston \etal(1999)]{Johnston99a}
  Johnston,~K.~V., Zhao,~H., Spergel,~D.~N., \& Hernquist,~L. 1999, \apj, 512, L109
\bibitem[Koposov \etal(2010)]{Koposov10a}
  Koposov,~S.~E., Rix,~H.-W., \& Hogg,~D.~W. 2010, \apj, 712, 260
\bibitem[McGill \& Binney(1990)]{McGill90a}
  McGill,~C. \& Binney,~J. 1990, \mnras, 244, 634
\bibitem[McMillan \& Binney(2008)]{McMillan08a}
  McMillan,~P.~J. \& Binney,~J.~J. 2008, \mnras, 390, 429
\bibitem[Ngan \& Carlberg(2014)]{Ngan14a}
  Ngan,~W.~H.~W. \& Carlberg,~R.~G. 2014, \apj, submitted
\bibitem[Pe{\~n}arrubia \etal(2012)]{Penarrubia12a}
  Pe{\~n}arrubia, J., Koposov,~S.~E., \& Walker,~M.~G. 2012, \apj, 760, 2
\bibitem[Price-Whelan \& Johnston(2013)]{PriceWhelan13a}
  Price-Whelan,~A.~M. \& Johnston,~K.~V.~2013, \apj, 778, L12
\bibitem[Sanders(2012)]{Sanders12a}
  Sanders,~J. 2012, \mnras, 426, 128
\bibitem[Sanders \& Binney(2013a)]{Sanders13a}
  Sanders,~J. \& Binney,~J. 2013a, \mnras, 433, 1813
\bibitem[Sanders \& Binney(2013b)]{Sanders13b}
  Sanders,~J. \& Binney,~J. 2013b, \mnras, 433, 1826
\bibitem[Tremaine(1999)]{Tremaine99a}
  Tremaine,~S. 1999, \mnras, 307, 877
\bibitem[Yoon \etal(2011)]{Yoon11a}
  Yoon,~J.~H., Johnston,~K.~V., \& Hogg,~D.~W. 2011, \apj, 731, 58
\end{thebibliography}

@@

\begin{figure}[t!]
 \includegraphics[width=0.48\textwidth,clip=]{aAI_jr.ps}
 \includegraphics[width=0.48\textwidth,clip=]{aAI_araz.ps}
  \caption{}\label{fig:aAI}
%python plot_aAI.py ../tex/aAI_jr.ps
%python plot_aAI.py ../tex/aAI_araz.ps
\end{figure}
